---
title: "week6"
output: html_document
---
```{r}
date()
```

```{r}

library(tidyverse)
library(dplyr)
library(ggplot2)

```

# Chapter 6: Analysis of longitudinal data

The Analysis exercise focuses on exploring your data and performing and interpreting simple straightforward methods (based on summary measures of data) and more advanced methods (called linear mixed effects models). For completing the Analysis exercises, include all the codes, your interpretations and explanations in the R Markdown file chapter6.Rmd.

Note: As you read the data sets in your chapter6.Rmd you may have to factor the categorical variables again, as R may not recognise them automatically as factors.


## 6.1 RATS

"Implement the analyses of Chapter 8 of MABS using the RATS data. (0-7 points: 0-4 points for graphs or analysis results + 0-3 points for their interpretations)"


### The data

The RATS data is from a nutrition study conducted in three groups of rats
(Crowder and Hand, 1990). The groups were put on different diets, and
each rats body weight (grams) was recorded repeatedly over a 9-week
period. (MABS, p .174) The variable "Time" = days.

```{r}

# read the RATS data
RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", header = TRUE, sep = '\t')

# Factor variables ID and Group
RATS$ID <- factor(RATS$ID)
RATS$Group <- factor(RATS$Group)

# Glimpse the data
glimpse(RATS)

# Convert data to long form
RATSL <- RATS %>%
  gather(key = WD, value = Weight, -ID, -Group) %>%
  mutate(Time = as.integer(substr(WD,3,4))) 

# Glimpse the data
glimpse(RATSL)


```

### RATS plots

```{r}

# The dimensions of the data
dim(RATSL)

# The individual rat growth profiles (non-standardized)
ggplot(RATSL, aes(x = Time, y = Weight, group = ID)) +
  geom_line(aes(linetype = Group)) +
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 10)) +
  scale_y_continuous(name = "Weight (grams)") +
  theme(legend.position = "top")

colnames(RATSL)
View(RATSL)

# Standardising the variable RATSL
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(std_weight = (Weight - mean(Weight))/sd(Weight)) %>%
  ungroup()

# Glimpse the data
glimpse(RATSL)
View(RATSL)

# The standardized RATSL weight plot by group, time and ID
ggplot(RATSL, aes(x = Time, y = std_weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=11)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "standardized weight")

# The individual rat growth profiles (standardized)
ggplot(RATSL, aes(x = Time, y = std_weight, group = ID)) +
  geom_line(aes(linetype = Group)) +
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 10)) +
  scale_y_continuous(name = "Weight (grams)") +
  theme(legend.position = "top")


```

###  Analysis 

Creating, and plotting the summary data. The RATS data does not seem to contain outliers, so there is no need to exclude them. There are 11 measurement points, for the 

```{r}


# Number of measurement days, baseline (day 1) included
n <- RATSL$Time %>% unique() %>% length()
n # =11

# Summary data with mean and standard error of weight grouped by Group and Time 
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise(mean = mean(Weight), se = sd(Weight)/sqrt(n) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSS)
View(RATSS)

# Plot the mean profiles
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.9,0.5)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")

# A boxplot of the mean versus group
ggplot(RATSS, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(weight), days")


```
### The analysis

The results of the ANOVA indicate that the differences between the groups were statistically significant. (I couldn't figure out how to add the baseline variable to the RATSL8S?).

Posthoc test?

```{r}

# Creating a summary data by Group and ID with mean as the summary variable (ignoring baseline time 1).
RATSL8S <- RATSL %>%
  filter(Time > "1") %>%
  group_by(Group) %>%
  summarise( Time = Time, 
             mean=mean(Weight) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSL8S)
View(RATSL8S)

# Add the baseline from the original data as a new variable to the summary data
View(RATS)

RATS2 <- RATS %>%
  gather(key = WD, value = Weight, -ID, -Group) 
View(RATS2)

RATSL8S <- RATSL8S %>%
  mutate(baseline = RATS2$WD1)

glimpse(RATSL8S)
View(RATSL8S)

# Fit the linear model with the mean as the response 
fit <- lm(mean ~ Group + Time, data = RATSL8S)

# Compute the analysis of variance table for the fitted model with anova()
anova(fit)

```


## 6.2 BPRS

"Implement the analyses of Chapter 9 of MABS using the BPRS data. (0-8 points: 0-4 points for graphs or analysis results + 0-4 points for their interpretations)"

The data consists of 40 male subjects, that were randomly assigned to one of two treatment groups and each subject was rated on the brief psychiatric rating scale (BPRS) measured before treatment began (week 0), and then at weekly intervals for eight weeks. The BPRS assesses the level of 18 symptom constructs, with scale 1 (not present) - 7 (extremely severe). (MABS, p. 157)

### The data

```{r}

# Read the BPRS data
BPRS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt", sep  =" ", header = T)

# Look at the (column) names and structure of BPRS
names(BPRS)
str(BPRS)

BPRS$treatment <- factor(BPRS$treatment)
BPRS$subject <- factor(BPRS$subject)

# Adding individual IDs
BPRS$ID <- seq.int(nrow(BPRS))
glimpse(BPRS)

# Converting to long form
BPRSL <-  BPRS %>% gather(key = weeks, value = bprs, -treatment, -subject, -ID)
glimpse(BPRSL)

# Extract the week number
BPRSL <-  BPRSL %>% mutate(week = as.integer(substr(weeks,5,5)))

# Take a glimpse at the BPRSL data
glimpse(BPRSL)
View(BPRSL)

# Plot of the bprs values by ID's and measurement times (weeks)
ggplot(BPRSL, aes(x = week, y = bprs, group = ID)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "weeks", breaks = seq(0,8)) +
  scale_y_continuous(name = "bprs") +
  theme(legend.position = "top")


```
### Standardizing the data

n <- 20

```{r}
# Standardise the variable bprs
BPRSL <- BPRSL %>%
  group_by(week) %>%
  mutate(stdbprs = (bprs - mean(bprs))/sd(bprs) ) %>%
  ungroup()

# Glimpse the data
glimpse(BPRSL)

```


### Plots


```{r}
# The distributions of bprs values
ggplot(BPRSL, aes(x= bprs)) + 
  facet_wrap("treatment", scales = "free") + 
  geom_bar()

# Plots with non-standardized values

ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))


# Plot with the standardized bprs

ggplot(BPRSL, aes(x = week, y = stdbprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  scale_y_continuous(name = "standardized bprs")


```

### Detecting and filtering the outlier

```{r}
BPRSL8S <- BPRSL %>%
  filter(week > 0) %>%
  group_by(treatment, subject, ID) %>%
  summarise( mean=mean(bprs) ) %>%
  ungroup()

# Glimpse the data
glimpse(BPRSL8S)

# Draw a boxplot of the mean versus treatment
ggplot(BPRSL8S, aes(x = treatment, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(bprs), weeks 1-8")

# Create a new data by filtering the outlier and adjust the ggplot code the draw the plot again with the new data
BPRSL8S1 <- BPRSL8S %>%
  filter(mean < 60)

# Draw a boxplot 2 of the mean versus treatment
ggplot(BPRSL8S1, aes(x = treatment, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(bprs), weeks 1-8")

# Glimpse the data
glimpse(BPRSL8S1)


```


## 6.3 Linear Mixed Effects Models 

"The main objective in the analysis of data from a longitudinal study is to
characterize change in the repeated values of the response variable and to determine the explanatory variables most associated with any change." (MABS, p. 169)

Multilevel models (also known as hierarchical linear models, linear mixed-effect model, mixed models, nested data models, random coefficient, random-effects models, random parameter models, or split-plot designs) are statistical models of parameters that vary at more than one level.They can be used as an alternative to ANCOVA (analysis of covariance), and they are able to analyze  experiments without the assumptions of homogeneity-of-regression slopes (that is required by ANCOVA)." [https://en.wikipedia.org/wiki/Multilevel_model]

"Linear mixed effects models are suitable for responses that can be assumed to be approximately normally distributed after conditioning on the explanatory variables." (MABS, p. 170)


### A regression model BPRSL_reg

Ignoring the repeated-measures structure of the data, a multiple linear regression model is fitted with bprs as dependent and week and treatment as explanatory variables. This model assumes independence of the repeated measures of bprs, and this assumption is highly unlikely.

The results of the model indicate, that the measurement time (week) was statistically significant explanatory variable (p<.000) for bprs scores, as the treatment group was not statistically significant (p=.661). The model explains 18% of the variation in bprs scores. 

```{r}

BPRSL_reg <- lm(bprs ~ week + treatment, data = BPRSL)

# print out a summary of the model
summary(BPRSL_reg)

```

### The Random Intercept Model

To begin the multilevel analysis of the BPRS data, first the random intercept model for the two explanatory variables (week & treatment) is fitted. Fitting a random intercept model allows the linear regression fit for each subject to differ in intercept from other subjects.

The first argument of lmer() function of the lme4 package is the formula object describing both the fixed-effects and random effects part of the model, with the response on the left of a ~ operator and the terms, separated by + operators, on the right. The random-effects terms distinguished by vertical bars (|).

In this model (bprs ~ week + treatment + (1 | ID)), where the week and treatment group are the fixed effects, compared to the multiple regression model, that was conducted first:

a) the standard error of the week is smaller (indicating that assuming independence will lead to the standard error of a within-subject covariate such as measurement time being larger than it should be because of ignoring the likely within-subject dependences, which will reduce the error variance in the model), and the t-value larger (incidating greater significance)
b) the the standard error of treatment group is larger, and the t-value is smaller (indicating that the significance is non-significant also in this model and even lower than in the first model)


```{r}
library(lme4)
library(Rcpp)

# Create a random intercept model

BPRSL_ref <- lmer(bprs ~ week + treatment + (1 | ID), data = BPRSL, REML = F)

# Print the summary of the model
summary(BPRSL_ref)


```

### A random intercept and random slope model

Fitting the random intercept and random slope model to the BPRS data allows the linear regression fits for each individual to differ in intercept but also in slope. This way it is possible to account for the individual differences in the bprs score profiles, but also the effect of measurement times and treatment.

The random intercept and slope model seems to provide a better ﬁt for these data: the AIC is slightly smaller, and in comparing the models (BPRSL_ref: bprs ~ week + treatment + (1 | ID), BPRSL_ref1: bprs ~ week + treatment + (week | ID)) with ANOVA  the chi-squared statistic of the latter is 63.663 (df 2), with a very small p-value (1.499e-14 ***).


```{r}

BPRSL_ref1 <- lmer(bprs ~ week + treatment + (week | ID), data = BPRSL, REML = FALSE)

# print a summary of the model
summary(BPRSL_ref1)

# perform an ANOVA test on the two models
anova(BPRSL_ref, BPRSL_ref1)

```

### Random Intercept and Random Slope Model with interaction

Finally, the ﬁt a random intercept and slope model is fitted. This model allows for a
group × time interaction. 

The fixed effect week*treatments t-value (1,349) is not significant. 
Comparison of the models (BPRSL_ref1: bprs ~ week + treatment + (week | ID), BPRSL_ref2: bprs ~ week * treatment + (week | ID)) shows, that there is no significant difference between them (chi-squared 1.78, p=0.1821). The AIC and the BIC are slightly larger in the model with interaction, so it is concluded that the random intercept and random slope model without the interaction provides a better model for the data. 


```{r}

# create a random intercept and random slope model
BPRSL_ref2 <- lmer(bprs ~ week * treatment + (week | ID), data = BPRSL, REML = FALSE)

# print a summary of the model
summary(BPRSL_ref2)

# perform an ANOVA test on the two models
anova(BPRSL_ref2, BPRSL_ref1)

# 
ggplot(BPRSL, aes(x = week, y = bprs, group = ID)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8)) +
  scale_y_continuous(name = "Bprs score") +
  theme(legend.position = "top")

# Create a vector of the fitted values
Fitted <- fitted(BPRSL_ref2)

# Create a new column fitted to RATSL
BPRSL <- BPRSL %>%
  mutate(Fitted)

# draw the plot of RATSL
ggplot(BPRSL, aes(x = week, y = Fitted, group = ID)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8)) +
  scale_y_continuous(name = "Bprs score") +
  theme(legend.position = "top")
 

```

